---
title: "Heart Attack Risk Prediction: Case Study"
author: "George Urgiles"
date: "Spring 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Load Libraries

```{r load-libraries}
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(janitor)
library(corrplot)
library(ggfortify)
```

# Load & Clean Data

```{r load-data}
set.seed(123)
data <- read.csv("Heart_Attack_Risk_Levels_Dataset.csv")
data <- data %>% clean_names()
data <- data %>%
  mutate(
    gender = factor(gender, labels = c("Male", "Female")),
    risk_level = as.factor(risk_level)
  )
glimpse(data)
sum(is.na(data))
```

# Descriptive Statistics

```{r summary-stats}
data %>% summary()
```

# Correlation Matrix for Numeric Features

```{r correlation-matrix}
numeric_data <- data %>% select(where(is.numeric))
corrplot(cor(numeric_data), method = "color", tl.cex = 0.7)
```

# Boxplots for Key Variables

```{r boxplots}
ggplot(data, aes(x = risk_level, y = troponin)) +
  geom_boxplot() +
  labs(title = "Troponin Levels by Risk Level")

ggplot(data, aes(x = risk_level, y = ck_mb)) +
  geom_boxplot() +
  labs(title = "CK-MB Levels by Risk Level")
```

# Explore Class Imbalance

```{r class-imbalance}
table(data$risk_level)
prop.table(table(data$risk_level))
```

## A pca plot a scatter plot used to visualize high-dimensional data by reducing it to two dimensions, typically using the first two principal components
##information accured from https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html

```{r pca-plot}
pca_data <- prcomp(numeric_data, scale. = TRUE)
autoplot(pca_data, data = data, colour = 'risk_level')
```

# Train/Test Split

```{r train-test-split}
train_index <- createDataPartition(data$risk_level, p = 0.7, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]
```

# Logistic Regression

```{r logistic-regression}
log_model <- train(
  risk_level ~ ., data = train_data,
  method = "multinom",
  trControl = trainControl(method = "cv", number = 5)
)
log_pred <- predict(log_model, test_data)
confusionMatrix(log_pred, test_data$risk_level)
```

# Decision Tree

```{r decision-tree}
tree_model <- rpart(risk_level ~ ., data = train_data, method = "class")
rpart.plot(tree_model, type = 2, extra = 104, fallen.leaves = TRUE)
tree_pred <- predict(tree_model, test_data, type = "class")
confusionMatrix(tree_pred, test_data$risk_level)
```

# Random Forest

```{r random-forest}
rf_model <- randomForest(risk_level ~ ., data = train_data, importance = TRUE, ntree = 500)
rf_pred <- predict(rf_model, test_data)
confusionMatrix(rf_pred, test_data$risk_level)
varImpPlot(rf_model)
```

# Model Comparison

```{r model-comparison}
results <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    confusionMatrix(log_pred, test_data$risk_level)$overall['Accuracy'],
    confusionMatrix(tree_pred, test_data$risk_level)$overall['Accuracy'],
    confusionMatrix(rf_pred, test_data$risk_level)$overall['Accuracy']
  )
)
print(results)
```

# Conclusion

We explored and modeled heart attack risk levels using logistic regression, decision trees, and random forests. We enhanced exploration with correlation plots, boxplots, and PCA to visualize data distribution. Random forests typically performed best, underlining the importance of key biomarkers like Troponin and CK-MB. Future work can include advanced tuning or deeper medical insights.
